Equalizer Programming Guide

### 2007-12-17 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and scalability page
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.5 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}

The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:

@@

    eq::Pipe*  pipe        = getPipe();
    Window*    firstWindow = static_cast< Window* >( pipe->getWindow( 0 ));

@@

    if( firstWindow == this )
    {
        _state = new VertexBufferState( getGLFunctions( ));
        _loadLogo();

@@

        if( initData.useVBOs() )
        {
            const eq::GLFunctions* glFunctions = getGLFunctions();
            // Check if all VBO funcs available, else leave DISPLAY_LIST_MODE on
            if( glFunctions->hasGenBuffers() && glFunctions->hasBindBuffer() &&
                glFunctions->hasBufferData() && glFunctions->hasDeleteBuffers())
            {
                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
                EQINFO << "VBO rendering enabled" << endl;
            }
            else
                EQWARN << "VBO function pointers missing, using display lists" 
                       << endl;
        }
    }
    else
    {
        _state       = firstWindow->_state;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_state ) // happens if first window failed to initialize
        return false;

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.6 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

\section{Scalable Rendering}

Real-time visualization is an inherently parallel
problem. Unfortunately, different applications have different rendering
algorithms, which require different scalable rendering modes to address
the right bottlenecks. Equalizer supports the most important algorithms,
and will continue to add new ones over time to meet application
requirements.

This section gives an introduction to scalable rendering, in order to
provide some background for application developers. The scalability
modes offered by Equalizer are discussed, along with their advantages
and disadvantages.

Choosing the right mode for the application profile is critical for
performance. Equalizer uses the concept of compounds to describe the
task decomposition and result recomposition. It allows the combination
of the different compound modes in any possible way, which allows the
address different bottlenecks in a flexible way.

\subsection{2D or Sort-First Compounds}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{\small A 2D compound}}
\end{wrapfigure}
2D or sort-first decomposes the rendering in screen-space, that is, each
contributing rendering unit processes a tile of the final view. The
recomposition simply assembles the tiles side-by-side on the destination
view.

The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the number of pixels of the destination view.

It's disadvantage is that it relies on view frustum culling to reduce
the amount of data rendered. Depending on the application data
structure, the overlap of some primitives between individual tiles
limits the scalability of this mode, typically to around eight graphics
cards. Each node has to potentially hold the full database for rendering.

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if possible.

\subsection{DB or Sort-Last Compounds}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\small A database compound}}
\end{wrapfigure}
DB or sort-last\footnote{3D model courtesy of AVS, USA.} decomposes the
rendered database so that all rendering units process a part of the
scene in parallel.

Volume rendering applications use an ordered alpha-based blending to
composite the result image. The depth buffer information is used to
composite the individual images correctly for polygonal data.

This mode provides very good scalability, since each rendering unit
processes a part of the database. This allows to lower the requirements
on all parts of the rendering pipeline: main memory usage, IO bandwidth,
GPU memory usage, vertex processing and fill rate.

Unfortunately, the database recomposition has linear increasing IO
requirements for the pixel transfer. Parallel recomposition algorithms,
such as direct-send address this problem by keeping the per-node IO
constant (see \fig{fDirectSend}).

The application has to partion the database so that the rendering units
render only part of the database. Some OpenGL features, do not work
correctly (anti-aliasing) or need special attention (transparency).

The best use of database compounds is to divide the data to a managable
size, and then to use other decomposition modes to achieve more
scalability. Volume rendering is one of the applications which can
profit from database compounds.

\subsection{Stereo Compounds}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\small A stereo compound}}
\end{wrapfigure}
Stereo compounds\footnote{3D model courtesy of Stereolithography Archive
  at Clemson University.} assign each eye pass to individual rendering
units. The resulting images are copied to the appropriate stereo
buffer. This modes supports a variety of stereo modes, including active
(quad-buffered) stereo, anaglyphic stereo and auto-stereo displays with
multiple eye passes.

Due to the frame consistency between the eye views, this modes scales
very well. The IO requirements for pixel transfer are small and
constant.

The number of rendering resources used by stereo compounds is limited by
the number of eye passes, typically two.

Stereo compounds are used by all applications when rendering in stereo,
and is often combined with other modes.

\subsection{Pixel Compounds}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\small A pixel compound}}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. The frustra of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels.

As 2D compounds, pixels compounds have low, constant IO requirements for
the pixel transfers during recomposition.

Pixel compounds only work very well for purely fill-limited
applications, techniques like frustum culling do not reduce the rendered
data for the source rendering resources.

OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple.

Pixel compounds are ideal for raytracing, which is highly fill-limited
and needs the full database for rendering anyway. Volume rendering
applications should choose this mode over 2D compounds.

@@

The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\vspace{-4ex}\end{wrapfigure}
The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:
\vspace{4ex}

@@

    eq::Pipe*  pipe        = getPipe();
    Window*    firstWindow = static_cast< Window* >( pipe->getWindows()[0] );

@@

    if( firstWindow == this )
    {
        _state = new VertexBufferState( glewGetContext( ));

@@

        if( initData.useVBOs() )
        {
            // Check if VBO funcs available, else leave DISPLAY_LIST_MODE on
            if( GLEW_VERSION_1_5 )
            {
                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
                EQINFO << "VBO rendering enabled" << endl;
            }
            else
                EQWARN << "VBO function pointers missing, using display lists" 
                       << endl;
        }

        if( initData.useGLSL() )
        {
            // Check if all functions are available
            if( GLEW_VERSION_2_0 )
            {
                EQINFO << "Shaders supported, attempting to load" << endl;
                _loadShaders();
            }
            else
                EQWARN << "Shader function pointers missing, using fixed "
                       << "pipeline" << endl;
        }

        _loadLogo();
    }
    else
    {
        _state       = firstWindow->_state;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_state ) // happens if first window failed to initialize
    {
        setErrorMessage( "No state handler object" );
        return false;
    }

    // turn off OpenGL lighting if we are using our own shaders
    if( _state->getProgram( pipe ) != VertexBufferState::FAILED )
        glDisable( GL_LIGHTING );

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.
